---
title: "Adding geographic features"
---

```{r}
#| warning: false
#| message: false

library(tidyverse)

library(rnaturalearth)
library(sf)
library(tidyverse)

sf_use_s2(FALSE)

worldmap = ne_countries(scale = 'medium', returnclass = 'sf')
```

A map can consist of many layers of geographic information stacked on top of each other.

As well as the basic countries and coastline maps, it is also possible to download lots of other specific geographic data from RNaturalearth, using a function `ne_download`. To use this, you'll need to specify the `category` and `type` of the map, as well as the `scale` and `returnclass` as before. The full list of categories and types is available on [this page](https://cran.r-project.org/web/packages/rnaturalearth/vignettes/rnaturalearth.html), towards the end.

The following example will download the 'rivers_lake_centerlines' data from the 'physical' category of maps:

```{r}

rivers = rnaturalearth::ne_download(scale = 'medium',category = 'physical', type = 'rivers_lake_centerlines', returnclass = 'sf')

```

We can add this river data using a new layer, with another `geom_sf` and a `+`.

One thing to note is that you always need to put the `coord_sf()` after your last geom_sf, otherwise it will ignore it and give the full map:

```{r}

worldmap |> 
  ggplot() +
  geom_sf() + 
  geom_sf(data = rivers) +  
  coord_sf(xlim = c(-10, 25), ylim = c(38, 60))


```

We can also change aspects of the map such as the fill of the shapes, and the colour of the outline. As with before, this is done using `color =` and `fill =`. To change all of the geographic data of a layer to a single color, simply type it within the `geom_sf` . If it is to be mapped to specific data, it should be enclosed within `aes()`

```{r}

worldmap |>
  ggplot() + 
  geom_sf(color = 'black', fill = 'lightgreen') +    
  geom_sf(data = rivers, color = 'blue') +    
  coord_sf(xlim = c(-10, 25), ylim = c(38, 60))


```
