---
title: "Opening and subsetting data"
---

### Learning objectives

-   Loading data from csv p

-   How to load the relevant packages

-   How to explore and adjust rows and columns (select, slice, arrange, rename)

    -   various select tools, starts_with, ends_with, contains.

    -   Mini exercises throughout

-   Arrange columns (and using -)

-   Arrange by multiple columns

-   Know how to change data types

-   How to read a csv file (local and internet)

-   How to write out to a file.

In the real world, we'll usually work with data which comes from spreadsheets and is tabular - meaning it consists of rows and columns. Later, we'll work with a real dataset from a spreadsheet, but to demonstrate the functions we'll use one built in from a package. You don't need to worry about this part of the code.

```{r}
titanic_df = titanic::titanic_train
```

The Tidyverse

We'll do all our data analysis and visualisation using a suite of tools called the Tidyverse. There are other ways to work with data in R, but this one works well for our purposes. To load the package, use:\

```{r}
library(tidyverse)
```

A function

A function is a set piece of code which does somehting in particular. For most of the functions we are using, the first thing we enter is the dataset, then further instructions depending on the function.

Select

We can use the function select() to select certain columns in our data. There are numerous instructions we can give to select() to do this. The simplest way is to give the column names, e.g.

```{r}

select(titanic_df, Name, Age, Sex)


```

We can also use position, e.g.

```{r}
select(titanic_df, 4,5,6)
```

Or a range of numbers, using \`:\`

```{r}
select(titanic_df, 1:4)
```

There are also helpers, starts_with(), ends_with(), contains():

```{r}

select(titanic_df, starts_with("S"))

```

```{r}
select(titanic_df, ends_with("ed"))
```

```{r}
select(titanic_df, contains("ar"))
```

You can also deselect columns by putting a minus sign (-) on front of the column name or range:

```{r}
select(titanic_df, -5)
```

It works with all the different select methods:

```{r}
select(titanic_df, -Age)
```

```{r}
select(titanic_df, -starts_with("S"))
```

If you want to deselect a range, you need to give the range as a vector, e.g.

```{r}
select(titanic_df, -c(1:3))
```

Slice

The other thing we might want to do is to pick particular rows. This is done using slice(). To pick a certain row, we can use:

```{r}
slice(titanic_df, 10)
```

You can also pick a range of numbers, as with select:

```{r}
slice(titanic_df, 1:10)
```

There are some other ways to pick rows, using some variants of slice. For example, you can pick rows based on the highest or lowest numbers of some variable. To do this, you can use slice_min() or slice_max(). In this case, you need to provide extra instructions. First the dataframe, as before, then the column you want to use, and finally the number of rows you would to return. The following code will return the ten oldest passengers on the titanic:

```{r}
slice_max(titanic_df, order_by = Age, n = 10)
```

What happens if you use slice_max or slice_min on a string?

```{r}
slice_max(titanic_df, order_by = Name, n = 20)
```

Do you notice something weird with the names? Why do you think that is?

Another basic operation is to arrange rows in a particular order. This can be done using the function arrange().

```{r}
arrange(titanic_df, Age)
```

You'll see that the data is sorted by the Age column, in ascending order (the smallest first). You can sort in descending order by putting a minus (-) sign before the column name:

```{r}

arrange(titanic_df, -Age)

```

You can also sort by multiple columns:

```{r}
arrange(titanic_df, Sex, -Age)
```

You'll notice here we sorted by a text column (Sex). The results are in alphabetical order, so this sorts all the female passengers by age, then all the male.

Exercises
