---
title: "Tmap"
---

## Interactive maps with tmap

Another feature to improve your data visualisations is to use interactive maps. Again, there are several ways to do this, including the package [Leaflet](https://rstudio.github.io/leaflet/).

The easiest way to create a map is with a package called tmap. This takes an sf object (similar to those we created in earlier weeks), and allows you to make a nice map using a syntax fairly similar to ggplot. Optionally, you can make these maps interactive.

To demonstrate this, I'll use a dataset from the [Shakespeare & Co project at Princeton](https://shakespeareandco.princeton.edu/), which contains information on the locations of members of the famous Shakespeare & Co bookshop and lending library in Paris.

First, install and load the tmap package.

```{r}
library(tmap)
library(tidyverse)
```

Load the dataset:

```{r}

sco_data = read_csv('SCoData_members_v1.2_2022_01.csv')

sco_data = sco_data %>% separate(coordinates, into = c('latitude', 'longitude'), sep = ',')

```

Now, create an sf object, as we learned in previous weeks

```{r}
library(sf)

sco_data_sf = st_as_sf(sco_data, coords = c('longitude', 'latitude'), na.fail = FALSE)

```

To specify an interactive map with tmap use `tmap_mode("view")`

```{r}
tmap_mode("view")
```

Build the map. First add a 'basemap', which is a zoomable world map. A full list of options is available [here](https://leaflet-extras.github.io/leaflet-providers/preview/).

Now, add the data using `tm_shape()`, and specify how it should be drawn using `tm_dots()`. The map below is interactive and clickable:

```{r}
# tm_basemap('OpenStreetMap.Mapnik') + 
#   tm_shape(sco_data_sf) + 
#   tm_dots()
```

## Extra: interactive maps with tmap.

To make nice interactive maps we'll use another package called tmap. `tmap` is a very nice package specifically for making maps. Unfortunately it doesn't use exactly the same syntax we have learned with ggplot, but it is by far the easiest way to make interactive maps.

The following code only works in RStudio so it is static in the book:

-   Load the tmap library.

-   Download a map from R Natural Earth

-   Switch to the 'view' mode using `tmap_mode("view")` to make the output an interactive map.

-   `tm_shape()` is more or less the equivalent of `ggplot()`: it's the 'base' of the map, where we specify which shapefile should be used. On top of this we add layers, such as polygons or points, and also additional map elements such as a legend, scale, projection, and so forth. We can also add additional `tm_shape()` layers.

-   Add the `tm_polygons()` layer which will draw polygon shapes on the map based on the shapefile. Set the `col` to the `income_grp` column. Note the difference from ggplot - where we would set the `fill`, and the column name would *not* be in quotation marks. Set the `id` which will determine which column is used to display text with a mouse hover.

-   Add a `tm_view()` layer, which allows us to set options for the interactive viewer. Add a vector of three numbers: the starting latitude and longitude, and the zoom level.

```{r}
#| error: false
#| warning: false
#| message: false

library(tmap)
library(tidyverse)
library(rnaturalearth)
library(sf)

sf_use_s2(FALSE)

worldmap = ne_countries(scale = 'medium', returnclass = 'sf') 

tmap_mode("view")

tm_shape(worldmap) +
  tm_polygons(fill="income_grp", id = 'name' ) +
  tm_view(set.view = c(0, 50, 3))

```

Tmap can do cool things like interactive facets, using the layer `tm_facets()`:

```{r}
tmap_mode("view")

tm_shape(worldmap) +
  tm_polygons(c("income_grp", 'pop_est'), id = 'name') +
  tm_view(set.view = c(0, 50, 4)) + 
  tm_facets(sync = TRUE, ncol = 2)
```
